---
title: "Stochastic Enrollment, Follow-up and Trial Calendar"
author: "Colin Longhurst"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: true
vignette: >
  %\VignetteIndexEntry{A User Guide for the `endpoints` Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r,echo=FALSE,warning=FALSE,message=FALSE}
pkgload::load_all(".")
library(tidyverse)
```

# Introduction

In many clinical trial simulations, the endpoint model is only one part of the data-generating process. Real trials also unfold over calendar time: subjects enroll gradually, follow-up differs between early and late enrollees, and final analysis may occur at a fixed calendar date, after the last randomized subject has reached a minimum amount of follow-up, or after a target number of events has accrued. This vignette introduces the stochastic enrollment, follow-up, and trial-calendar features in `{endpoints}`. We show how to simulate homogeneous Poisson-process enrollment, piecewise Poisson-process enrollment with changing accrual rates, subject-level follow-up limits, and several trial-ending rules. These tools are especially useful for time-to-event simulations, where enrollment timing and administrative censoring directly affect observed event times and censoring indicators.


# Stochastic enrollment with `enrollment_details`

The `enrollment_details` argument controls how subjects enter the trial calendar. If no stochastic enrollment is specified, all subjects are treated as enrolling at calendar time 0. When stochastic enrollment is used, `makeData()` generates an `enrollTime` column, where `enrollTime` is the subject's calendar enrollment time **relative** to the first randomized subject.

In `{endpoints}`, stochastic enrollment is modeled using a Poisson-process accrual framework. In the simplest case, enrollment follows a homogeneous Poisson process with constant accrual rate \(\lambda\). Equivalently, the waiting times between consecutively enrolled subjects are exponentially distributed:

\[
G_i \sim \mathrm{Exponential}(\lambda),
\]

and enrollment times are generated as cumulative sums:

\[
T_{E,i} = \sum_{m=1}^{i} G_m.
\]

The first enrollment time is shifted to 0, so the trial calendar starts at the first randomized subject.

A more flexible option is piecewise Poisson-process accrual. In this setting, the accrual rate changes over calendar intervals. If the current trial calendar time lies in interval \(k\), then the waiting time to the next enrolled subject is generated using that interval's rate:

\[
G \sim \mathrm{Exponential}(\lambda_k).
\]

If the proposed next enrollment crosses a cutpoint, the process advances to that cutpoint and continues using the next interval's rate. This allows users to represent common trial accrual patterns such as slow start-up, ramp-up, plateau enrollment, or temporary enrollment pauses.

For this section, we use a single continuous endpoint so that we can focus entirely on enrollment behavior.


## Homogeneous Poisson-process enrollment

Here we demonstrate the simple case when `simulate_distribution = "exponential"`. In this example, we assume a two-arm trial with 300 subjects per arm and a single continuous endpoint. We set `enrollment_exponential_rate = 10` meaning that the trial enrolls, on average, 10 subjects per unit of calendar time.



```{r homogeneous-poisson-sim}
ep1 <- list(
  endpoint_type = "continuous",
  baseline_mean = 10,
  sd = 5,
  trt_effect =  -2
)

sim_exp <- makeData(
  correlation_matrix = NULL,
  sample_size_per_group = 300,
  SEED = 101,
  endpoint_details = list(ep1),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 10
  )
  )

dat_exp <- as.data.frame(sim_exp)
```

When enrollment_details() is included, the resulting dataframe includes a `enrollTime` column:

```{r}
dat_exp %>% head() %>% knitr::kable(.)
```

The constant enrollment rate can be visualized using a histogram or density plot: 

```{r homogeneous-poisson-hist, fig.width=7, fig.height=4}
ggplot2::ggplot(dat_exp, aes(x = enrollTime)) +
  geom_histogram(bins = 35,fill = "steelblue", alpha = 0.35,
    color = "black",linewidth = 0.3) +
  labs(
    x = "Enrollment time",
    y ="Number enrolled",
    title = "Homogeneous Poisson-process enrollment",) +
  theme_bw()
```

Because the rate is constant, the slope of the cumulative enrollment curve is also constant: 
```{r}
enroll_curve_exp <- dat_exp %>%
  arrange(enrollTime) %>%
  mutate(cumulative_n = row_number())

ggplot2::ggplot(enroll_curve_exp, aes(x = enrollTime, y = cumulative_n)) +
  geom_step(color = "steelblue", linewidth = 0.7) +
  #geom_vline(
  #  xintercept = cutpoints_piece[-c(1, length(cutpoints_piece))],
  #  linetype = "dashed",
  #  color = "firebrick"
  #) +
  labs(
    x = "Enrollment time",
    y = "Cumulative number enrolled",
    title = "Cumulative enrollment under exponential accrual"
  ) +
  theme_bw()

```


## Piecewise Poisson-process enrollment

The `piecewise` option allows the accrual rate to change across pre-specified
calendar intervals. This is useful when the expected enrollment pattern is not
constant over time, for example if recruitment starts slowly, accelerates as
sites open, and then levels off once the trial is fully active.

In the example below, we simulate a two-arm trial with 200 subjects per arm
and the same continuous endpoint as above. We specify four enrollment
intervals:

- calendar times 0 to 12 with rate 5 subjects per unit time,
- calendar times 12 to 24 with rate 10 subjects per unit time,
- calendar times 24 to 36 with rate 15 subjects per unit time, and
- calendar times 36 to 52 with rate 20 subjects per unit time.

```{r piecewise-poisson-sim}
sim_piece <- makeData(
  correlation_matrix = NULL,
  sample_size_per_group = 500,
  SEED = 203,
  endpoint_details = list(ep1),
  enrollment_details = list(
    enrollment_distribution        = "piecewise",
    piecewise_enrollment_cutpoints = c(0, 12, 24, 36, 52),
    piecewise_enrollment_rates     = c(5, 10, 15, 20)
  )
)

dat_piece <- as.data.frame(sim_piece)

cutpoints_piece <- c(0, 12, 24, 36, 52)
rates_piece <- c(5, 10, 15, 20)
interval_labels_piece <- c("[0,12)", "[12,24)", "[24,36)", "[36,52)", "[52,Inf)")
tail_length_piece <- max(dat_piece$enrollTime) - tail(cutpoints_piece, 1)

interval_summary_piece <- tibble(
  enroll_interval = factor(interval_labels_piece, levels = interval_labels_piece),
  interval_length = c(diff(cutpoints_piece), NA_real_),
  rate_expected = c(rates_piece, tail(rates_piece, 1))
) %>%
  left_join(
    dat_piece %>%
      mutate(
        enroll_interval = cut(
          enrollTime,
          breaks = c(cutpoints_piece, Inf),
          right = FALSE,
          include.lowest = TRUE,
          labels = interval_labels_piece
        )
      ) %>%
      count(enroll_interval, name = "n_enrolled", .drop = FALSE),
    by = "enroll_interval"
  ) %>%
  mutate(
    n_expected = interval_length * rate_expected,
    n_expected = if_else(
      is.na(interval_length),
      nrow(dat_piece) - sum(n_expected, na.rm = TRUE),
      n_expected
    ),
    rate_observed = case_when(
      is.na(interval_length) & n_enrolled == 0 ~ 0,
      is.na(interval_length) ~ n_enrolled / tail_length_piece,
      TRUE ~ n_enrolled / interval_length
    )
  ) %>%
  select(
    enroll_interval,
    n_enrolled,
    n_expected,
    rate_observed,
    rate_expected
  )

enroll_curve_piece <- dat_piece %>%
  arrange(enrollTime) %>%
  mutate(cumulative_n = row_number())
```

The table below shows the realized number of enrolled subjects in each calendar
interval. Because enrollment is stochastic, the observed counts will vary from
run to run even when the target piecewise rates are fixed. The final row represents
the remaining subjects enrolled after the last user-specified cutpoint. **Please note that the last piecewise enrollment rate continues beyond the final cutpoint until all planned subjects have enrolled.**

```{r piecewise-poisson-table}
interval_summary_piece %>%
  knitr::kable(
    caption = "Observed enrollment by calendar interval"
  )
```

<!-- If users want to choose piecewise accrual rates to target specific enrollment -->
<!-- proportions by interval, the rates can be obtained by simple algebra. Suppose -->
<!-- the interval widths are \(d_1, d_2, d_3, d_4\), and the desired proportions are -->
<!-- \(p_1 = 0.15\), \(p_2 = 0.35\), \(p_3 = 0.25\), and \(p_4 = 0.25\). Then the -->
<!-- first three rates can be solved sequentially as -->
<!-- \[ -->
<!-- \lambda_1 = -\frac{\log(1-p_1)}{d_1}, \qquad -->
<!-- \lambda_2 = -\frac{\log\left(1-\frac{p_2}{1-p_1}\right)}{d_2}, \qquad -->
<!-- \lambda_3 = -\frac{\log\left(1-\frac{p_3}{1-p_1-p_2}\right)}{d_3}. -->
<!-- \] -->
<!-- These formulas come from writing each interval probability as the probability -->
<!-- of still being unenrolled at the start of the interval times the probability of -->
<!-- enrolling inside that interval. For the final interval, note that the current -->
<!-- implementation continues enrollment beyond the last cutpoint using the final -->
<!-- rate. Therefore, if the target proportions sum to 100% by the end of the last -->
<!-- finite interval, the last rate would formally need to be infinite to force all -->
<!-- remaining subjects to enroll before that cutpoint. In practice, a common -->
<!-- approach is to solve the early-interval rates exactly, then choose a suitably -->
<!-- large final rate so that most of the remaining subjects accrue in the last -->
<!-- interval. -->

The histogram below shows how the distribution of enrollment times changes as
the accrual rate increases across intervals.

```{r piecewise-poisson-hist, fig.width=7, fig.height=4}
ggplot2::ggplot(dat_piece, aes(x = enrollTime)) +
  geom_histogram(
    binwidth = 2,
    boundary = 0,
    fill = "steelblue",
    alpha = 0.35,
    color = "black",
    linewidth = 0.3
  ) +
  geom_vline(
    xintercept = cutpoints_piece[-c(1, length(cutpoints_piece))],
    linetype = "dashed",
    color = "firebrick"
  ) +
  labs(
    x = "Enrollment time",
    y = "Number enrolled",
    title = "Piecewise Poisson-process enrollment"
  ) +
  theme_bw()
```

Finally, the cumulative enrollment curve makes the changing accrual pattern
especially clear. The slope becomes steeper in later intervals because the
piecewise enrollment rate is larger.

```{r piecewise-poisson-cumulative, fig.width=7, fig.height=4}
ggplot2::ggplot(enroll_curve_piece, aes(x = enrollTime, y = cumulative_n)) +
  geom_step(color = "steelblue", linewidth = 0.7) +
  geom_vline(
    xintercept = cutpoints_piece[-c(1, length(cutpoints_piece))],
    linetype = "dashed",
    color = "firebrick"
  ) +
  labs(
    x = "Enrollment time",
    y = "Cumulative number enrolled",
    title = "Cumulative enrollment under piecewise accrual"
  ) +
  theme_bw()
```

# Patient Follow-Up and Trial Duration with `followup_details` and `trial_end_details`

The `followup_details` argument controls subject-level follow-up limits after
enrollment. It has two fields:

- `min_followup`: a planned minimum follow-up time, mainly used by
  trial-ending rules such as `"last_patient_min_followup"` or
  `require_min_followup = TRUE` in event-driven trials.
- `max_followup`: a hard cap on how much follow-up any one subject can
  contribute.

Conceptually, `{endpoints}` first determines each subject's calendar
enrollment time, then determines the overall trial end time, and finally
computes each subject's available follow-up as:

\[
\mathrm{availableFollowup}_i =
\min\left(\mathrm{max\_followup}, \, \mathrm{trial\_end\_time} - T_{E,i}\right),
\]

with truncation at 0 if a subject enrolls after the effective trial end. If
`max_followup` is omitted, follow-up is limited only by the trial end. If
`min_followup` is omitted, it does not constrain when the trial is allowed to
end.

A simple example is:

```{r}
followup_details <- list(
  min_followup = 6,
  max_followup = 18
)
```

This means the design aims for the last randomized subject to have at least 6
time units of follow-up before final analysis, while no individual subject can
contribute more than 18 time units of follow-up.

The `trial_end_details` argument controls when the trial stops on the calendar
time scale. In `{endpoints}`, this can be handled in several ways:

- `type = "none"`: no calendar-based trial end is imposed.
- `type = "fixed_calendar"`: the trial ends at a user-specified calendar time.
- `type = "last_patient_min_followup"`: the trial ends once the last
  randomized subject has reached `min_followup`.
- `type = "event_driven"`: the trial ends when a target number of events is
  observed for a specified time-to-event endpoint.

For event-driven designs, `trial_end_details` also allows the user to specify
which endpoint drives stopping (`event_endpoint`), how many events are required
(`target_events`), whether a minimum follow-up constraint must also be
respected (`require_min_followup`), and what should happen if the event target
is not reached (`target_not_reached` together with an optional
`max_trial_duration`).

In particular, `require_min_followup = TRUE` is useful when the trial should
not stop immediately at the moment the event target is reached. Instead, the
final analysis is delayed until both conditions are satisfied: the required
number of events has occurred and the last randomized subject has reached
`min_followup`. This is a common way to combine an event-driven stopping rule
with a design requirement that late enrollees still contribute a minimum amount
of follow-up.

In other words, `followup_details` determines how much follow-up each subject
can contribute, while `trial_end_details` determines when the trial is allowed
to stop. These two settings work together to define the final
`availableFollowup` for each enrolled subject.

## Example: fixed calendar trial end

The simplest trial-ending rule is a fixed calendar cutoff. In this setting, the
user directly specifies the final analysis time through
`trial_end_details$trial_end_time`, and each subject's follow-up is then
determined by how early or late they entered the study.

To illustrate, consider a two-arm time-to-event trial with 100 subjects per
arm. Suppose the annual event rate in the control group is 30%, the treatment
hazard ratio is 0.6, and there is also 15% random censoring. We measure
time in months, so we convert the annual event probability into a monthly
exponential hazard and use `rate_from_prob()` to obtain an approximate random
censoring rate.

```{r tte-fixed-setup}
baseline_rate_fixed <- rate_from_prob(
  target_prob = 0.30,
  mode = "admin",
  admin_time = 12
)

censor_rate_fixed <- rate_from_prob(
  target_prob = 0.85,
  mode = "simple",
  event_rate = baseline_rate_fixed
)

ep_tte_fixed <- list(
  endpoint_type = "tte",
  baseline_rate = baseline_rate_fixed,
  trt_effect = log(0.6),
  censoring_rate = censor_rate_fixed,
  fatal_event = TRUE
)
```

In the example below, we assume moderate exponential enrollment and set a fixed
calendar trial end at 30 months. 

```{r tte-fixed-calendar}
sim_tte_fixed <- makeData(
  correlation_matrix = NULL,
  sample_size_per_group = 100,
  SEED = 41,
  endpoint_details = list(ep_tte_fixed),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 12
  ),
  trial_end_details = list(
    type = "fixed_calendar",
    trial_end_time = 30
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

dat_tte_fixed <- as.data.frame(sim_tte_fixed)
tc_tte_fixed <- sim_tte_fixed$meta$trial_calendar

tte_fixed_summary <- tibble(
  trial_end_time = tc_tte_fixed$trial_end_time,
  max_enroll_time = max(dat_tte_fixed$enrollTime),
  mean_available_followup = mean(dat_tte_fixed$availableFollowup),
  observed_events = sum(dat_tte_fixed$Status_1)
)

tte_fixed_summary %>%
  knitr::kable(
    digits = 2,
    caption = "Fixed calendar trial end"
  )
```

Because the trial always ends at calendar time 30, earlier enrollees have more
potential follow-up than later enrollees. In this design, the resulting
follow-up is approximately
\[
\mathrm{availableFollowup}_i =
\min( 30 - T_{E,i}),
\]
truncated at 0 if a subject enrolls after the fixed calendar cutoff.

```{r tte-fixed-head}
dat_tte_fixed %>%
  select(trt, TTE_1, Status_1, enrollTime, availableFollowup) %>%
  arrange(enrollTime) %>% 
  head() %>%
  knitr::kable(
    digits = 2,
    caption = "First six patients enrolled showing demonstrating fixed follow-up time"
  )
```

## Example: last-patient-minimum-follow-up design

We now move to a slightly more design-driven rule, where the trial stays open
until the last randomized subject has reached a minimum amount of follow-up. To
illustrate these ideas, consider a two-arm time-to-event trial with 100
subjects per arm. Suppose the annual event rate in the control group is 30%,
the treatment hazard ratio is 0.6, and there is also 15% random censoring. We
measure time in months, so we first convert the annual event rate into a
monthly exponential hazard. We then use `followup_details` to require at least
24 months of follow-up for the last randomized subject, while capping any one
subject's follow-up at 36 months.

```{r tte-followup-setup}
baseline_rate_monthly <- rate_from_prob(
  target_prob = 0.30,
  mode = "admin",
  admin_time = 12
)

censor_rate_monthly <- rate_from_prob(
  target_prob = 0.85,
  mode = "simple",
  event_rate = baseline_rate_monthly
)

ep_tte_followup <- list(
  endpoint_type = "tte",
  baseline_rate = baseline_rate_monthly,
  trt_effect = log(0.6),
  censoring_rate = censor_rate_monthly,
  fatal_event = TRUE
)
```

In the first version, we assume a moderate exponential enrollment rate of 15
subjects per month across the trial.

```{r tte-followup-moderate}
sim_tte_mod <- makeData(
  correlation_matrix = NULL,
  sample_size_per_group = 100,
  SEED = 42,
  endpoint_details = list(ep_tte_followup),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 15
  ),
  followup_details = list(
    min_followup = 24,
    max_followup = 36
  ),
  trial_end_details = list(
    type = "last_patient_min_followup"
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

dat_tte_mod <- as.data.frame(sim_tte_mod)
tc_tte_mod <- sim_tte_mod$meta$trial_calendar

tte_mod_summary <- tibble(
  scenario = "Moderate enrollment",
  enrollment_rate = 15,
  trial_end_time = tc_tte_mod$trial_end_time,
  max_enroll_time = max(dat_tte_mod$enrollTime),
  mean_available_followup = mean(dat_tte_mod$availableFollowup),
  observed_events = sum(dat_tte_mod$Status_1)
)

tte_mod_summary %>%
  knitr::kable(
    digits = 2,
    caption = "Moderate enrollment with last-patient minimum follow-up"
  )
```

Because the trial ends when the last randomized subject has 24 months of
follow-up, earlier enrollees contribute more than 24 months, up to the maximum
cap of 36 months. The resulting dataset includes both `enrollTime` and
`availableFollowup`:

```{r tte-followup-head}
dat_tte_mod %>%
  select(trt, TTE_1, Status_1, enrollTime, availableFollowup) %>%
  head() %>%
  knitr::kable(
    digits = 2,
    caption = "Example rows from the moderate-enrollment simulation"
  )
```

Now suppose we change only the enrollment speed, making accrual slower at 8
subjects per month while keeping all endpoint and follow-up settings fixed.

```{r tte-followup-slower}
sim_tte_slow <- makeData(
  correlation_matrix = NULL,
  sample_size_per_group = 100,
  SEED = 42,
  endpoint_details = list(ep_tte_followup),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 8
  ),
  followup_details = list(
    min_followup = 24,
    max_followup = 36
  ),
  trial_end_details = list(
    type = "last_patient_min_followup"
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

dat_tte_slow <- as.data.frame(sim_tte_slow)
tc_tte_slow <- sim_tte_slow$meta$trial_calendar

tte_compare <- tibble(
  scenario = c("Moderate enrollment", "Slower enrollment"),
  enrollment_rate = c(15, 8),
  trial_end_time = c(tc_tte_mod$trial_end_time, tc_tte_slow$trial_end_time),
  max_enroll_time = c(max(dat_tte_mod$enrollTime), max(dat_tte_slow$enrollTime)),
  mean_available_followup = c(
    mean(dat_tte_mod$availableFollowup),
    mean(dat_tte_slow$availableFollowup)
  ),
  observed_events = c(sum(dat_tte_mod$Status_1), sum(dat_tte_slow$Status_1))
)

tte_compare %>%
  knitr::kable(
    digits = 2,
    caption = "Slower enrollment produces more observed follow-up and events"
  )
```

In this paired comparison, the slower enrollment scenario has a later last
patient in, so the overall trial lasts longer on the calendar scale. That
gives earlier subjects more opportunity to reach the 36-month follow-up cap,
which in turn increases the average available follow-up and the observed number
of events.

## Example: event-driven trial with a fatal stopping endpoint

We next consider an event-driven design with two time-to-event endpoints. The
first endpoint is fatal and is used to trigger the stopping rule. The second
endpoint is non-fatal and is included to illustrate how the overall trial
duration can affect the number of secondary events observed by the time the
fatal endpoint reaches its event target.

For this example, we use a modest positive dependence between the two TTE
endpoints, an annual fatal event rate of 20% in the control group, and an
annual non-fatal event rate of 40%. We use hazard ratios of 0.75 for the fatal
endpoint and 0.85 for the non-fatal endpoint. To keep the example realistic,
we also include independent random censoring for both endpoints.

```{r event-driven-setup}
cor_tte2 <- corr_make(
  num_endpoints = 2,
  values = rbind(c(1, 2, 0.25))
)

fatal_rate_monthly <- rate_from_prob(
  target_prob = 0.20,
  mode = "admin",
  admin_time = 12
)

nonfatal_rate_monthly <- rate_from_prob(
  target_prob = 0.40,
  mode = "admin",
  admin_time = 12
)

ep_fatal <- list(
  endpoint_type = "tte",
  baseline_rate = fatal_rate_monthly,
  trt_effect = log(0.75),
  censoring_rate = 0.01,
  fatal_event = TRUE
)

ep_nonfatal <- list(
  endpoint_type = "tte",
  baseline_rate = nonfatal_rate_monthly,
  trt_effect = log(0.85),
  censoring_rate = 0.015,
  fatal_event = FALSE
)
```

In the first scenario, enrollment follows a simple homogeneous Poisson process
with exponential inter-arrival times. We stop the trial once 70 fatal events
have been observed on `TTE_1`, and we cap individual follow-up at 48 months.

```{r event-driven-exponential}
sim_event_exp <- makeData(
  correlation_matrix = cor_tte2,
  sample_size_per_group = 150,
  SEED = 77,
  endpoint_details = list(ep_fatal, ep_nonfatal),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 12
  ),
  followup_details = list(
    max_followup = 48
  ),
  trial_end_details = list(
    type = "event_driven",
    event_endpoint = "TTE_1",
    target_events = 70
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

dat_event_exp <- as.data.frame(sim_event_exp)
tc_event_exp <- sim_event_exp$meta$trial_calendar

event_exp_summary <- tibble(
  scenario = "Exponential enrollment",
  trial_end_time = tc_event_exp$trial_end_time,
  fatal_events = sum(dat_event_exp$Status_1),
  nonfatal_events = sum(dat_event_exp$Status_2),
  mean_available_followup = mean(dat_event_exp$availableFollowup)
)

event_exp_summary %>%
  knitr::kable(
    digits = 2,
    caption = "Event-driven trial with exponential enrollment"
  )
```

The realized calendar quantities are stored in `sim_event_exp$meta$trial_calendar`,
which includes the trial end time, the stopping reason, whether the event target
was reached, and the number of events observed at the realized trial end.

Now we hold the endpoint and stopping-rule settings fixed, but replace the
simple exponential accrual model with a piecewise enrollment pattern that
starts more slowly and accelerates later in the trial.

```{r event-driven-piecewise}
sim_event_piece <- makeData(
  correlation_matrix = cor_tte2,
  sample_size_per_group = 150,
  SEED = 77,
  endpoint_details = list(ep_fatal, ep_nonfatal),
  enrollment_details = list(
    enrollment_distribution = "piecewise",
    piecewise_enrollment_cutpoints = c(0, 12, 24, 36),
    piecewise_enrollment_rates = c(3, 8, 15)
  ),
  followup_details = list(
    max_followup = 48
  ),
  trial_end_details = list(
    type = "event_driven",
    event_endpoint = "TTE_1",
    target_events = 70
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

dat_event_piece <- as.data.frame(sim_event_piece)
tc_event_piece <- sim_event_piece$meta$trial_calendar

event_compare <- tibble(
  scenario = c("Exponential enrollment", "Piecewise enrollment"),
  trial_end_time = c(
    tc_event_exp$trial_end_time,
    tc_event_piece$trial_end_time
  ),
  fatal_events = c(
    sum(dat_event_exp$Status_1),
    sum(dat_event_piece$Status_1)
  ),
  nonfatal_events = c(
    sum(dat_event_exp$Status_2),
    sum(dat_event_piece$Status_2)
  ),
  mean_available_followup = c(
    mean(dat_event_exp$availableFollowup),
    mean(dat_event_piece$availableFollowup)
  )
)

event_compare %>%
  knitr::kable(
    digits = 2,
    caption = "Recruitment pattern affects event-driven trial duration and secondary events"
  )
```

In both scenarios, the stopping rule is driven by the fatal endpoint, so the
number of observed fatal events is similar by construction. However, the
piecewise accrual pattern stretches the trial over a longer calendar duration,
which gives more time for secondary non-fatal events to accrue before the
fatal endpoint reaches its target. This is one of the main reasons it can be
important to model enrollment explicitly when simulating event-driven designs.

## Event-driven stopping with `require_min_followup = TRUE`

In some event-driven trials, reaching the target number of events is not
sufficient by itself. Investigators may also want the last randomized subject
to contribute a minimum amount of follow-up before the final analysis. This can
be enforced by setting `require_min_followup = TRUE` together with
`followup_details$min_followup`.

```{r event-driven-min-followup}
sim_event_no_min <- makeData(
  correlation_matrix = cor_tte2,
  sample_size_per_group = 150,
  SEED = 77,
  endpoint_details = list(ep_fatal, ep_nonfatal),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 12
  ),
  followup_details = list(
    min_followup = 12,
    max_followup = 48
  ),
  trial_end_details = list(
    type = "event_driven",
    event_endpoint = "TTE_1",
    target_events = 70,
    require_min_followup = FALSE
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

sim_event_with_min <- makeData(
  correlation_matrix = cor_tte2,
  sample_size_per_group = 150,
  SEED = 77,
  endpoint_details = list(ep_fatal, ep_nonfatal),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 12
  ),
  followup_details = list(
    min_followup = 12,
    max_followup = 48
  ),
  trial_end_details = list(
    type = "event_driven",
    event_endpoint = "TTE_1",
    target_events = 70,
    require_min_followup = TRUE
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

dat_event_no_min <- as.data.frame(sim_event_no_min)
dat_event_with_min <- as.data.frame(sim_event_with_min)

tc_event_no_min <- sim_event_no_min$meta$trial_calendar
tc_event_with_min <- sim_event_with_min$meta$trial_calendar

event_min_compare <- tibble(
  scenario = c("No minimum follow-up", "Require minimum follow-up"),
  trial_end_time = c(
    tc_event_no_min$trial_end_time,
    tc_event_with_min$trial_end_time
  ),
  trial_end_reason = c(
    tc_event_no_min$trial_end_reason,
    tc_event_with_min$trial_end_reason
  ),
  n_events_at_end = c(
    tc_event_no_min$n_events_at_end,
    tc_event_with_min$n_events_at_end
  ),
  mean_available_followup = c(
    mean(dat_event_no_min$availableFollowup),
    mean(dat_event_with_min$availableFollowup)
  )
)

event_min_compare %>%
  knitr::kable(
    digits = 2,
    caption = "Minimum follow-up can delay an event-driven final analysis"
  )
```

When `require_min_followup = TRUE`, the trial does not stop immediately when
the fatal endpoint first reaches its event target. Instead, the final analysis
is delayed until the last randomized subject has also reached the required
minimum follow-up.

## If the event target is not reached

By default, an event-driven trial errors if the requested event target cannot
be reached under the simulated design. The `target_not_reached` argument allows
the user to choose a fallback rule instead. In the example below, we request an
intentionally unattainable fatal event target and ask the simulation to fall
back to `"last_patient_min_followup"`.

```{r event-driven-target-not-reached}
sim_event_fallback <- suppressWarnings(
  makeData(
    correlation_matrix = cor_tte2,
    sample_size_per_group = 80,
    SEED = 90,
    endpoint_details = list(ep_fatal, ep_nonfatal),
    enrollment_details = list(
      enrollment_distribution = "exponential",
      enrollment_exponential_rate = 12
    ),
    followup_details = list(
      min_followup = 12,
      max_followup = 24
    ),
    trial_end_details = list(
      type = "event_driven",
      event_endpoint = "TTE_1",
      target_events = 1000,
      target_not_reached = "last_patient_min_followup"
    ),
    non_fatal_censors_fatal = FALSE,
    target_correlation = FALSE
  )
)

dat_event_fallback <- as.data.frame(sim_event_fallback)
tc_event_fallback <- sim_event_fallback$meta$trial_calendar

event_fallback_summary <- tibble(
  trial_end_time = tc_event_fallback$trial_end_time,
  trial_end_reason = tc_event_fallback$trial_end_reason,
  event_target_reached = tc_event_fallback$event_target_reached,
  n_events_at_end = tc_event_fallback$n_events_at_end
)

event_fallback_summary %>%
  knitr::kable(
    digits = 2,
    caption = "Fallback behavior when the event target is not reached"
  )
```

Here the fatal event target is not reached, so the realized trial end is driven
by the fallback rule rather than by the event target itself. A similar approach
can be used with `target_not_reached = "max_trial_duration"` together with
`max_trial_duration`.

```{r event-driven-max-duration-snippet, eval=FALSE}
sim_event_max_duration <- makeData(
  correlation_matrix = cor_tte2,
  sample_size_per_group = 80,
  SEED = 90,
  endpoint_details = list(ep_fatal, ep_nonfatal),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 12
  ),
  followup_details = list(
    max_followup = 24
  ),
  trial_end_details = list(
    type = "event_driven",
    event_endpoint = "TTE_1",
    target_events = 1000,
    target_not_reached = "max_trial_duration",
    max_trial_duration = 30
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)
```

## Subjects enrolled after the realized trial end

In an event-driven design, the realized trial end time is not known in advance.
As a result, some planned subjects may be generated with enrollment times after
the final trial cutoff. These rows remain in the simulated dataset, but they
have `availableFollowup = 0` and therefore contribute no analyzable follow-up.

```{r event-driven-zero-followup}
sim_event_zero_fu <- makeData(
  correlation_matrix = cor_tte2,
  sample_size_per_group = 150,
  SEED = 77,
  endpoint_details = list(ep_fatal, ep_nonfatal),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 4
  ),
  followup_details = list(
    max_followup = 48
  ),
  trial_end_details = list(
    type = "event_driven",
    event_endpoint = "TTE_1",
    target_events = 30
  ),
  non_fatal_censors_fatal = FALSE,
  target_correlation = FALSE
)

dat_event_zero_fu <- as.data.frame(sim_event_zero_fu)

dat_event_zero_fu %>%
  filter(availableFollowup == 0) %>%
  select(trt, TTE_1, Status_1, TTE_2, Status_2, enrollTime, availableFollowup) %>%
  head() %>%
  knitr::kable(
    digits = 2,
    caption = "Subjects enrolled after the realized event-driven trial end"
  )
```

These rows are usually excluded from the analysis population. In practice, the
event-driven analysis set is often defined as:

```{r event-driven-analysis-pop}
dat_event_analysis <- dat_event_zero_fu %>%
  filter(availableFollowup > 0)
```

That is, subjects with `availableFollowup == 0` should typically be filtered
out before summarizing or analyzing the realized event-driven dataset.



