---
title: "Simulating Longitudinal Data with `endpoints`"
author: "Colin Longhurst"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: true
vignette: >
  %\VignetteIndexEntry{Simulating Longitudinal Data with `endpoints`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,warning=FALSE,message=FALSE}
library(endpoints)
library(tidyverse)
```

# Introduction


There are two ways in which the `endpoints` package can be used to generate longitudinal data. The first is by directly generating correlated outcomes and the second is by generating random random effects. Examples of each method can be found below.

# Correlated Responses

## Longitudinal Data Only

Suppose we want to simulate a longitudinal continuous outcome measured at a fixed set of visits, with trajectories that differ by treatment group. For illustration, consider the marginal mean model

$$
Y_{i t}=\beta_0+\beta_{\text {time }} t+\beta_{\text {trt }} x_i+\boldsymbol\beta_{\text {time:trt }} t x_i+\varepsilon_{i t}, \quad i=1, \ldots, n, \quad t \in\{0,1, \ldots, T-1\},
$$

where $x_i \in\{0,1\}$ is the treatment indicator and the residual vector $\varepsilon_i=\left(\varepsilon_{i 0}, \ldots, \varepsilon_{i, T-1}\right)^{\top}$ follows

$$
e_i \sim \mathcal{N}\left(\mathbf{0}, \sigma^2 \mathbf{V}\right), \quad \mathbf{V}_{t t^{\prime}}=\rho^{\left|t-t^{\prime}\right|}
$$


Here $\mathbf{V}$ is the standard AR(1) correlation structure: measurements closer in time are more strongly correlated. More explicitly, this can be written as 

 $$\mathbf{V}=\left(\begin{array}{ccccc}1 & \rho & \rho^2 & \cdots & \rho^{n-1} \\ \rho & 1 & \rho & \cdots & \rho^{n-2} \\ \rho^2 & \rho & 1 & \cdots & \rho^{n-3} \\ \vdots & \vdots & \vdots & \ddots & \vdots \\ \rho^{n-1} & \rho^{n-2} & \rho^{n-3} & \cdots & 1\end{array}\right)$$

In this example we use:

- $\mathrm{AR}(1)$ correlation $\rho=0.5$,
- $T=5$ visits (baseline + 4 follow-up visits),
- a constant control-group mean over time,
- a non-linear treatment effect across visits,
- residual SD $\sigma=0.95$.

**Step 1: Specify the within-subject correlation**

We encode the AR(1) structure directly using `corr_make()`:

```{r}
N_visits <- 5 
rho <- 0.5

corMat_5_visits <- corr_make(
  num_endpoints = N_visits,
  values = rbind(
    c(1,2,rho), c(1,3,rho^2), c(1,4,rho^3),c(1,5,rho^4),
    c(2,3,rho), c(2,4,rho^2), c(2,5,rho^3),
    c(3,4,rho), c(3,5,rho^2),
    c(4,5,rho)
  )
)
```

**Step 2: Treat each visit as a "correlated endpoint"**

Although `endpoint_details` typically describes distinct endpoints, we can also use it to represent distinct visits of a single longitudinal outcome. Each entry below specifies the marginal mean at a visit for control, plus the visit-specific treatment effect.

```{r}
sigma <- 0.95 # constant residual error

baseline <- list(
  endpoint_type = "continuous",
  baseline_mean = 0.45,
  trt_effect    = 0,     # no baseline difference
  sd            = sigma
)

v1 <- list(endpoint_type = "continuous", baseline_mean = 0.45, trt_effect = -0.10, sd = sigma)
v2 <- list(endpoint_type = "continuous", baseline_mean = 0.45, trt_effect = -0.15, sd = sigma)
v3 <- list(endpoint_type = "continuous", baseline_mean = 0.45, trt_effect = -0.25, sd = sigma)
v4 <- list(endpoint_type = "continuous", baseline_mean = 0.45, trt_effect = -0.40, sd = sigma)
```

**Step 3: Simulate wide data with makeData()**

```{r}
correlated_visits_wide <- makeData(
  correlation_matrix    = corMat_5_visits,
  SEED                  = 321,
  sample_size_per_group = 2000,   # large n to check asymptotic behavior
  endpoint_details      = list(baseline, v1, v2, v3, v4)
)
```


```{r}
knitr::kable(
  head(correlated_visits_wide$data),
  caption = "Wide data returned by makeData() (one column per visit)"
)
```

At this stage we have a standard "wide" format dataset: each visit appears as a separate column (`Cont_1`, …, `Cont_5`). The key observation is that a collection of correlated endpoints can be interpreted as repeated measures once we reshape the data.

**Step 4: Convert wide $\rightarrow$ long**

```{r,message=FALSE,warning=FALSE}
library(tidyverse)
correlated_visits_long <- 
  correlated_visits_wide$data %>%
  rowid_to_column('ID') %>% # keep track of patient
  gather(.,key="visit",value='Measure',-trt,-ID) %>%   # convert wide to long data
  mutate(ID = as.factor(ID),
         # clean up names
         visit  = case_match(visit,
                             "Cont_1" ~ 0,
                             "Cont_2" ~ 1,
                             "Cont_3" ~ 2,
                             "Cont_4" ~ 3,
                             "Cont_5" ~ 4)) %>%
  arrange(ID)
```

The long data now looks like a typical longitudinal dataset:
```{r}
knitr::kable(
  head(correlated_visits_long),
  caption = "Longitudinal data after reshaping to long format"
)
```

**Step 5: Validate the intended effects and correlation**

To verify that the simulated data reflect the intended mean structure and AR(1) correlation, we fit a marginal model using `geeM::geem()`:

```{r}
ge1 <- geeM::geem(
  Measure ~ factor(visit) * trt,
  id     = ID,
  corstr = "ar1",
  data   = correlated_visits_long,
  family = gaussian
)
```

Inspecting the output from the model, we observe that the simulation proceeded correctly:

```{r}
est_effects <- c(
  round(summary(ge1)$beta[7:10], 3),  # treatment effects at visits 1-4 (vs baseline)
  round(summary(ge1)$phi, 2),         # residual SD (see note below)
  round(summary(ge1)$alpha, 2)        # AR(1) correlation estimate
)

true_effects <- c(-0.10, -0.15, -0.25, -0.40, sigma, rho)

labels_latex <- c(
  "$\\beta_{\\mathrm{trt},1}$",
  "$\\beta_{\\mathrm{trt},2}$",
  "$\\beta_{\\mathrm{trt},3}$",
  "$\\beta_{\\mathrm{trt},4}$",
  "$\\sigma$",
  "$\\rho$"
)

out_tbl <- data.frame(
  Parameter = labels_latex,
  True      = true_effects,
  Estimated = est_effects
)

knitr::kable(
  out_tbl,
  escape = FALSE,
  digits = 3,
  caption = "True vs. estimated longitudinal treatment effects and correlation parameters."
)
```

(Note: $\sigma$ is the marginal residual SD in $\varepsilon_{it} \sim N(0,\sigma^2\mathbf{V})$)

We now have a longitudinal data set with the correct data characteristics to use in simulation. Finally, we can visualize a small set of subject trajectories:


```{r}
set.seed(888)
IDs = c(sample(1:4000,8)) 
correlated_visits_long %>%
  filter(.,ID %in% IDs) %>%
  ggplot(.,aes(x = visit,y=Measure,group = ID,color = as.factor(trt)))+
  geom_line() +
  geom_point() +
  labs(title = "",
       x = "Trial Time",
       y = "Measure",
       color = "Treatment") +
  theme_minimal()+
  scale_color_brewer(palette = "Set1")
```


## Longitudinal Data with Other Endpoints

To correlate longitudinal measurements with additional endpoints (e.g., a time-to-event outcome), we expand:

1. the correlation matrix to include the new endpoint(s), and
2. the `endpoint_details` list to include a TTE specification,

then reshape only the longitudinal columns while keeping the additional endpoint columns in place.

**Example: add a non-fatal hospitalization endpoint (TTE)**

Here we assume an exchangeable within-visit correlation for simplicity, and specify correlations between hospitalization time and each visit.


```{r}
rho <- 0.15

corMat_long_w_hosp <- corr_make(
  num_endpoints = 6,
  values = rbind(
    # exchangeable correlation among visits (1...5)
    c(1,2,rho), c(1,3,rho), c(1,4,rho), c(1,5,rho),
    c(2,3,rho), c(2,4,rho), c(2,5,rho),
    c(3,4,rho), c(3,5,rho),
    c(4,5,rho),

    # correlations between visits and hospitalization (endpoint 6)
    c(1,6,0.00), c(2,6,0.10), c(3,6,0.20), c(4,6,0.25), c(5,6,0.15)
  )
)
```

We specify hospitalization as a non-fatal TTE endpoint (no censoring here for simplicity):

```{r}
hosp <- list(
  endpoint_type  = "tte",
  baseline_rate  = 1/2,        
  trt_effect     = log(0.80),   
  fatal_event    = FALSE     
)
```

Simulate the data: 
```{r}
wide_w_hosp <- makeData(
  correlation_matrix    = corMat_long_w_hosp,
  SEED                  = 123,
  sample_size_per_group = 2000,
  endpoint_details      = list(baseline, v1, v2, v3, v4, hosp)
)
```

It is often helpful to verify the simulated marginals and (approximate) correlation structure before reshaping:

```{r,eval=F}
summary(wide_w_hosp)
```

<details>
<summary>Show summary(wide_w_hosp)</summary>

```{r}
#| echo: false
#| results: asis
x <- paste(capture.output(summary(wide_w_hosp)), collapse = "\n")
cat("<pre>", x, "</pre>", sep = "\n")
```
</details> 

Now reshape only the longitudinal columns, leaving the TTE endpoint intact:

```{r}
long_w_hosp <- wide_w_hosp$data %>%
  rowid_to_column("ID") %>%
  pivot_longer(
    cols = starts_with("Cont_"),
    names_to = "visit",
    values_to = "Measure"
  ) %>%
  mutate(
    ID = as.factor(ID),
    visit = dplyr::case_match(
      visit,
      "Cont_1" ~ 0,
      "Cont_2" ~ 1,
      "Cont_3" ~ 2,
      "Cont_4" ~ 3,
      "Cont_5" ~ 4
    )
  ) %>%
  arrange(ID, visit)

head(long_w_hosp, 10) %>% knitr::kable()
```

This produces a simulated data set with a TTE endpoint that is correlated with a longitudinal measurement:

# Random Effects

An alternative way to generate longitudinal data using the `endpoints` framework is to simulate subject-specific random effects and then use those latent quantities inside a separate longitudinal data-generating process. This approach is often more natural when the longitudinal outcome is intended to follow a mixed-effects model, as it directly represents patient-level heterogeneity through random intercepts and slopes.

For example, consider the following longitudinal model with a random intercept and random slope:

$$
\begin{gathered}
y_{ijt}
=
\underbrace{(\beta_0 + \beta_1 j + \alpha_i)}_{\text{intercepts}}
+
t\,\underbrace{(\beta_2 + \beta_{1:2} j + \gamma_i)}_{\text{slopes}}
+
\epsilon_{ijt}
\\
\epsilon_{ijt} \sim N(0,\sigma^2)
\quad \text{and} \quad
\begin{pmatrix}
\alpha_i \\
\gamma_i
\end{pmatrix}
\sim
\mathcal{N}
\!\left(
\begin{pmatrix}
0 \\
0
\end{pmatrix},
\mathbf{\Lambda}
\right) \quad \text{where} \quad\mathbf{\Lambda} =
\begin{pmatrix}
\sigma_\alpha^2 & \rho_{\alpha\gamma}\sigma_\alpha\sigma_\gamma \\
\rho_{\alpha\gamma}\sigma_\alpha\sigma_\gamma & \sigma_\gamma^2
\end{pmatrix}.
\end{gathered}
$$
Here:
- $j \in\{0,1\}$ indicates treatment group ( $0=$ control, $1=$ treatment),
- $i$ indexes subjects,
- $t$ indexes time,
- $\alpha_i$ is the subject-specific random intercept,
- $\gamma_i$ is the subject-specific random slope.

In this formulation, the correlation in repeated measurements is induced through the shared random effects rather than by directly specifying a within-subject correlation matrix across visits.

**Two-stage Simulation Strategy**

The basic idea is to proceed in two steps:

1. Use `makeData()` to simulate the subject-level latent variables (e.g., random intercepts, random slopes, and any additional endpoints such as TTE outcomes).

2. Use those subject-level quantities inside a second helper function that expands the data over time and generates the repeated measurements.

This approach is especially useful when the longitudinal outcome must be correlated with other endpoints (e.g., hospitalization or mortality), since `makeData()` can generate all of those latent subject-level quantities jointly.

**Example: Random Effects & Two TTE Endpoints**

In this example we simulate a trial with:

- one fatal TTE endpoint: all-cause mortality (ACM),
- one non-fatal TTE endpoint: hospitalization (HP),
- one latent random intercept,
- one latent random slope.

We first define the correlation matrix, which encodes the dependence among these four subject-level quantities:

```{r}
cor_rf <- corr_make(num_endpoints = 4,
                    values = rbind(
                      c(1,2,.2), # ACM ~ HP
                      c(1,3,.15), # ACM ~ random intercept
                      c(1,4,.1), # ACM ~ random slope
                      c(2,3,.1), # HP ~ random intercept
                      c(2,4,.05), # HP ~ random slope
                      c(3,4,.3) # random intercept ~ slope
                    )
)
```

Next, we specify the two random effects as continuous endpoints centered at 0:

```{r}
random_intercept <- list(
  endpoint_type = "continuous",
  baseline_mean = 0,
  sd            = 0.5
)

random_slope <- list(
  endpoint_type = "continuous",
  baseline_mean = 0,
  sd            = 0.25
)
```

We then define the two TTE endpoints:

```{r}
ACM <- list(
  endpoint_type  = "tte",
  baseline_rate  = 1/4,        
  trt_effect     = log(0.80),
  censoring_rate = 1/8,
  fatal_event    = TRUE     
)

HP <- list(
  endpoint_type  = "tte",
  baseline_rate  = 1/2,        
  trt_effect     = log(0.80),
  censoring_rate = 1/15,
  fatal_event    = FALSE     
)
```

Now we use `makeData()` to generate the subject-level dataset, where each subjects has their latent random effects generated:

```{r}
rf_data <- makeData(
  correlation_matrix = cor_rf,
  sample_size_per_group = 1000,
  SEED = 888,
  endpoint_details = list(ACM,HP,random_intercept,random_slope),
  non_fatal_censors_fatal = TRUE,
  enrollment_details = list(
    administrative_censoring = 4
  )
)
```

At this stage, the simulated dataset contains:

- TTE_1: all-cause mortality,
- TTE_2: hospitalization,
- Cont_1: random intercept,
- Cont_2: random slope.

These are all subject-level quantities. The remaining step is to expand the data over time and generate the repeated longitudinal measurements.

**Creating the Longitudinal Data**

To do this, we use a small helper function that:

- expands each subject across a time grid,
- evaluates a user-supplied linear predictor,
- adds Gaussian residual error,
- optionally censors the longitudinal outcome after a fatal event.

```{r}
longMaker_lite <- function(data,
                           time_sequence,
                           linear_predictor,
                           residual_error,
                           censor_longitudinal_data = FALSE) {
  
  expanded_data <- data %>%
    rowid_to_column("ID") %>%
    tidyr::expand_grid(time = time_sequence) %>%
    mutate(linPred = eval(parse(text = linear_predictor)))
  
  expanded_data$response <- expanded_data$linPred +
    rnorm(nrow(expanded_data), mean = 0, sd = residual_error)
  
  # If requested, censor longitudinal measurements after the fatal TTE endpoint
  # (here using TTE_1 only, for illustration)
  if (censor_longitudinal_data) {
    expanded_data$response <- ifelse(
      expanded_data$time > expanded_data$TTE_1,
      NA,
      expanded_data$response
    )
  }
  
  expanded_data
}
```

In the example below, the linear predictor includes:

- a fixed intercept of 10,
- no baseline treatment effect,
- a control-group slope of $-\mathbf{0 . 5}$,
- a treatment-by-time interaction of +1 ,
- a subject-specific random intercept ( `Cont_1` ),
- a subject-specific random slope ( `Cont_2 * time` ).

We first generate the latent longitudinal outcome **without** censoring, so that we can verify the fixed and random effects separately from the TTE censoring mechanism.

```{r}
long_rf_data_test <- longMaker_lite(data = rf_data$data,
                                    time_sequence = seq(from=0,to = 2, by = 0.25),
                                    linear_predictor =
                                      "10 + (0*trt) + (-.5*time) + (1*trt*time) + Cont_1 + (Cont_2*time)",
                                    residual_error =1,
                                    censor_longitudinal_data = FALSE
)
```

**Validate the Data**

We can now fit a linear mixed model to verify that the simulated fixed effects, random effects, and induced correlation structure are behaving as intended:
```{r}
lme4::lmer(
  response ~ trt * time + (time | ID),
  data = long_rf_data_test
) %>%
  summary(correlation = FALSE)
```

The estimated treatment-by-time interaction, residual variation, and random-effects covariance are close to their data-generating values. 

Next, we regenerate the longitudinal data, this time allowing the fatal event (ACM:`TTE_1`) to censor future longitudinal measurements:

```{r}
long_rf_data_final <- longMaker_lite(data = rf_data$data,
                                     time_sequence = seq(from=0,to = 2, by = 0.25),
                                     linear_predictor =
                                       "10 + (0*trt) + (-.5*time) + (1*trt*time) + Cont_1 + (Cont_2*time)",
                                     residual_error =1,
                                     censor_longitudinal_data = TRUE)
```

This yields a dataset in which the repeated continuous outcome is jointly related to two TTE endpoints through the shared subject-level latent structure, and is additionally truncated after the fatal event.

```{r}
knitr::kable(head(long_rf_data_final,16))
```


# Comparing the approaches

| Approach | Advantages | Disadvantages |
|---|---|---|
| **Correlated endpoints** | - No separate data generating process<br>- Only requires correlation parameters (no random effects)<br>- Allows for complicated correlation patterns | - Must specify $K \times t$ correlation parameters in addition to the sub-longitudinal correlation matrix ($K$ is number of additional endpoints) |
| **Random effects** | - Far fewer correlation parameters to specify<br>- MMRM/mixed models are common in industry<br>- Flexible data generating process<br>- More explicit generative assumptions | - Requires a second data generating process |

























