---
title: "A User Guide for the `endpoints` Package"
author: "Colin Longhurst"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: true
vignette: >
  %\VignetteIndexEntry{A User Guide for the `endpoints` Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r,warning=FALSE,message=FALSE}
library(endpoints)
library(tidyverse)
```


# Introduction

The `makeData()` framework is designed to simulate flexible clinical trial datasets with a mix of endpoint types and trial-design features, while keeping the interface simple and modular. It supports both straightforward single-endpoint simulations and more complex multivariate settings with correlated outcomes generated through a Gaussian copula (NORTA-style) construction. The function also accommodates:

- trials with multiple treatment arms
- time-to-event outcomes with censoring (competing or semi-competing risks)
- realistic trial logistics such as stochastic enrollment and administrative follow-up

In this vignette, we illustrate the package through a sequence of examples, beginning with simple single-endpoint simulations and gradually introducing more advanced features.

# Quick Start: One Endpoint Simulations

The core function in `endpoints` is `makeData()`, which is used across all simulation settings. In the simplest case, `makeData()` can simulate a single-endpoint trial with one or more treatment arms.

For one-endpoint simulations, the most important arguments are:

- `sample_size_per_group`: Sample size per treatment arm (scalar or vector)
- `endpoint_details`: A list of endpoint specifications (one sub-list per endpoint)

An optional argument used in multivariate settings is:

- `correlation_matrix`: A correlation matrix describing dependence across endpoints

This is discussed in later sections.

```{r,eval=FALSE}
makeData(correlation_matrix = NULL,
  sample_size_per_group = 1000,
  endpoint_details = list(
    ...
    )
  )
```


## Specifying endpoints `endpoint_details`

The `makeData()` function allows for four endpoint types, which are encoded via the `endpoint_details` argument: continuous, binary, count and time-to-event. We discuss each below.

### Continuous

An example specification of a continuous endpoint is seen below:

```{r,eval=T}
c_ep1 <- list(
      endpoint_type = "continuous",
      baseline_mean = 10,
      sd            = c(3,2),
      trt_effect    = -2
    )
```

When generating continuous endpoints, `makeData()` uses a Gaussian (normal) marginal distribution. The endpoint specification arguments are:

- `baseline_mean`: Mean in the control group.
- `sd`: If a scalar is provided, a common SD is used across groups. If a vector is provided, it specifies arm-specific SDs.
- `trt_effect = NULL`: Mean difference(s) for treatment group(s) relative to control. A scalar is used for one treatment arm; a vector may be used for multiple treatment arms. If `NULL`, only control-group data are generated.

For a two-arm study, `trt_effect = -2` implies `treatment mean = baseline_mean - 2`

```{r}
simple1 <- makeData(correlation_matrix = NULL,
  sample_size_per_group = 1000,
  SEED = 1,
  endpoint_details = list(c_ep1)
  )
```

A quick summary of the simulation details is displayed when printing the resulting object:
```{r}
simple1
```

To confirm the distributions have the correct parameters, we can use the summary function

```{r}
knitr::kable(summary(simple1)$continuous)
```

From above, we see that the estimated baseline mean, treatment effect and standard deviation parameter are all close to the true values.

The resulting simulated data set can be accessed from the main `makeDataSim` object:

```{r}
head(simple1$data)
```

The response variable is automatically encoded as `Cont_1` to indicate that it is the first continuous endpoint.

### Binary

For binary outcomes, the event probability is modeled on the logit scale. You may specify treatment effects either directly on the probability scale (`trt_prob`) or on the log-odds scale (`trt_effect`), but not both.

```{r}
bin_ep1 <- list(
      endpoint_type  = "binary",
      baseline_prob  = 0.30,
      trt_prob = 0.45
    )
```

Binary endpoints are specified via the following arguments:

- `baseline_prob`: Probability of success in the control group.
- `trt_prob`: Scalar or vector of treatment-group success probabilities.
- `trt_effect = NULL`: Optional scalar or vector of treatment effects on the log-odds scale (relative to control). Only one of `trt_prob` or `trt_effect` may be supplied.

Once again we can confirm that the simulated data has the proper characteristics:

```{r}
simple2 <- makeData(correlation_matrix = NULL,
  sample_size_per_group = 1000,
  SEED = 2,
  endpoint_details = list(bin_ep1)
  )
knitr::kable(summary(simple2)$binary)
```

### Count

To generate count outcomes, we use the zero-inflated negative binomial distribution with a log-link function. The parameterization arguments are as follows:

```{r}
int_ep1 <- list(
      endpoint_type = "count",
      baseline_mean = 8,    
      trt_count = 10,
      size          = 100, # approximate Poisson       
      p_zero        = 0     
    )
```

- `baseline_mean`: Mean count in the control group.
- `trt_count`: Scalar or vector of treatment-group mean counts.
- `trt_effect = NULL`: Optional scalar or vector of treatment effects on the log rate-ratio scale. Only one of `trt_count` or `trt_effect` may be supplied.
- `size`: Negative binomial dispersion parameter (larger values reduce overdispersion and approach the Poisson limit). In the parameterization used, the variance of the distribution is equal to $\mu + \frac{\mu^2}{\phi}$. Here, `size` controls $\phi$. 
- `p_zero`: Zero-inflation probability.

An example use can be seen below:

```{r}
simple3 <- makeData(correlation_matrix = NULL,
  sample_size_per_group = 1000,
  SEED = 2,
  endpoint_details = list(int_ep1)
  )
knitr::kable(summary(simple3)$count)
```


*Note:* When `size` is large, the negative binomial approaches a Poisson distribution. In those cases, the dispersion parameter may be difficult to estimate precisely from a single simulated dataset (e.g., via `MASS::glm.nb()`), which is usually not a practical concern for simulation.


### Time-to-Event Endpoints

Time-to-event endpoints are supported through exponential event-time models with optional independent censoring, fatal/non-fatal event logic, and the trial-calendar interface built from `enrollment_details`, `followup_details`, and `trial_end_details`. We return to these features later in this vignette and in the dedicated `enrollment_guide` vignette.

# Multiple Treatment Arms

To extend the above simulations to settings with several treatment groups, we can simply replace scalar inputs with vectors. In general, if there are $G$ total study arms (including control), then treatment-specific vectors (e.g., `trt_prob`, `trt_effect`, `sd`) should typically have length $G-1$, corresponding to the non-control arms in order. 

For example, imagine we have a 4-grp study with a binary endpoint:

```{r}
bin_4grp <- list(
      endpoint_type  = "binary",
      baseline_prob  = 0.30,
      trt_prob = c(0.35, 0.40, 0.45)
    )

sim_bin4grp <- makeData(correlation_matrix = NULL,
  sample_size_per_group = c(100,200,400,400),
  SEED = 123,
  endpoint_details = list(bin_4grp)
  )
sim_bin4grp
```

Once again we can ensure the distributions are being simulated correctly by calling `summary`:

```{r}
knitr::kable(summary(sim_bin4grp)$binary)
```



# Correlated Endpoints with Gaussian Copula

Gaussian copulas provide a flexible way to generate multivariate outcomes with user-specified marginal distributions and dependence. Let $\mathbf{Z} = (Z_1,\dots,Z_J)^\top \sim \mathrm{MVN}(\mathbf{0}, \Sigma_Z)$. Applying the probability integral transform componentwise gives

$$
X_j = F_j^{-1}\{\Phi(Z_j)\}, \qquad j=1,\dots,J,
$$

where $\Phi$ is the standard normal CDF and $F_j^{-1}$ is the inverse CDF (quantile function) of the desired marginal distribution for endpoint $j$. This construction preserves the chosen marginals and induces dependence across endpoints through the Gaussian copula.

Because Pearson correlation is generally **not** preserved under nonlinear monotone transformations, `makeData()` optionally uses numerical calibration (`target_correlation = TRUE`) to find a latent Gaussian correlation matrix that approximately reproduces the requested Pearson correlation matrix on the observed endpoint scale. 

**Important:** When `target_correlation = FALSE`, the supplied `correlation_matrix` is used directly as the latent Gaussian correlation matrix (the copula parameter), not as a calibrated Pearson correlation matrix on the observed endpoint scale. It is recommended that this option is set to `TRUE`. See Cario & Nelson (1997) for more details.


## Creating the Correlation Matrix

To create a correlation matrix, we use the `corr_make()` function which has two arguments:

- `num_endpoints`: The number of correlated endpoints to generate
- `values`: Pairwise correlations specified as a 3-column object (e.g., matrix or data frame), where each row has the form `(i, j, rho)` giving the correlation between endpoints `i` and `j`. Unspecified pairs default to a correlation value of 0. 

Consider a trial with 3 endpoints with the following correlation:

```{r}
corMat_3 <- corr_make(
  num_endpoints =   3,
  values = rbind(
    c(1,2,.2), c(1,3,.1),
    c(2,3,.15)
  )
) 
```


## Simulating Correlated Endpoints

Using the endpoints defined above, we can now generate a trial with three correlated endpoints, one continuous, one binary and one count:

```{r}
sim_3eps <- makeData(
  correlation_matrix    = corMat_3,
  sample_size_per_group = 3000,
  SEED = 777,
  endpoint_details = list(
    c_ep1,
    bin_ep1,
    int_ep1
  )
)
```

To quickly confirm that the estimated correlation between endpoints is close to the correlation specified, we use the `summary` function:

<details>
<summary>Show summary(sim_3eps)</summary>

```{r}
#| echo: false
#| results: asis
x <- paste(capture.output(summary(sim_3eps)), collapse = "\n")
cat("<pre>", x, "</pre>", sep = "\n")
```
</details> 



From the summary above, we see that the within-arm correlation is close to that specified in `corMat_3` and that the marginal distributions have the correct distributional characteristics. Note that because the correlation calibration is numerical (and summaries are based on finite samples), small discrepancies between the requested and empirical within-arm correlations are expected.

Here we also introduce the `plot` function, which allows users to quickly confirm simulation details:

```{r}
plot(sim_3eps)
```

Note that different experimental arms can be plotted by toggling the `arm` argument, and that names may also be supplied:

```{r}
plot(sim_3eps,arm=1,names=c("BMI","Illness Worsened","# episodes"))
```


The code can easily be modified to accommodate several treatment groups by applying the conventions discussed in the previous section.

# Time-to-event Endpoints and Censoring

In `makeData()`, all time-to-event (TTE) endpoints are distributed as exponential random variables $f(x;\lambda) = \lambda e^{-\lambda x}$. They are encoded as elements in `endpoint_details` as follows:

- `baseline_rate`: The exponential event-rate parameter $\lambda$ for the control group (so the mean event time is $1/\lambda$).
- `trt_effect`: Scalar or vector of treatment effects on the log hazard-ratio scale.
- `censoring_rate = NULL`: Exponential censoring-rate parameter for independent censoring (if `NULL`, no random censoring is applied).
- `fatal_event`: If `TRUE`, the endpoint is treated as fatal and censors subsequent non-fatal TTE endpoints. See below for details.

```{r}
tte_ep1 <- list(
      endpoint_type  = "tte",
      baseline_rate  = 1/24,        
      trt_effect     = log(0.80),   
      censoring_rate = NULL,        
      fatal_event    = F     
    )
```

**Censoring**

For a time-to-event endpoint, for the $i^{th}$ patient we observe $X_i = \min\{T_i, C_i\}$ where $T \sim Exp(\lambda_e)$ is the event time and $C_i \sim Exp(\lambda_c)$ is the censoring time. For the primary terminal event (or a non-terminal event when there are no terminal events), the probability of observing an event is $$P({T}_i < {C}_i) = \int^\infty_0\lambda_ee^{-\lambda_ex}e^{-\lambda_cx} = \frac{\lambda_{e}}{\lambda_e+\lambda_c},$$

which can be used to select a censoring rate that produces the user-specified event rate. In the example from the code chunk above, if we want to select a censoring rate such that we observe an event rate of .9, some algebra yields $\lambda_c = \frac{1/24}{0.90} -\frac{1}{24} = 1/216$. This can be calculated by using the helper function `rate_from_prob()`:

```{r}
lambda_c <- rate_from_prob(
  target_prob = 0.90, # target proportion of patients with events
  mode = "simple", 
  event_rate = 1 / 24 # rate parameter for primary event
)
print(paste0("1","/",1/lambda_c))
```


When there are semi-competing risks, and the TTE endpoints have low correlation, the event rate for the secondary, non-fatal TTE endpoint can be approximated as $\frac{\lambda_{e2}}{\lambda_{e1}+\lambda_{c1}+\lambda_{e2}+\lambda_{c2}}$ where the subscripts $1, 2$ denote the first and secondary endpoints. The above formula is meant only as an approximation as the dependence (induced by the copula) between $X_1$ and $X_2$ is not accounted for. This approximation can also be handled by the helper function. In this case, the output is the rate of the censoring mechanism for the secondary endpoint. 

```{r}
rate_from_prob(
  target_prob = 0.20, # hypothetical example
  mode = "semi-competing",
  fatal_event_rate = 1 / 50,
  fatal_censor_rate = 1 / 16.667,
  nonfatal_event_rate = 1 / 35
)
```

Please see `?rate_from_prob()` for details.

We also note here that the censoring mechanism is independent of the copula so pairwise correlations involving censored TTE endpoints generally will not equal the specified coefficient. To confirm the simulation is working as intended, we recommend first simulating the data without censoring, confirming the correlation, then re-generating the data with censoring. 

**Fatal Events**

Let $X_1$ be a fatal event and $X_2$ be a non-fatal event, then we write $X_2 = \min\{T_1, C_1, T_2,C_2\}$. That is, we cannot observe any event that occurs after a fatal event occurs **or** is censored.

If a non-fatal event occurs before a fatal event ($T_2 < \{T_1, C_1,C_2\}$), a fatal event can still be subsequently observed (or censored). If a non-fatal event is censored before a fatal event ($C_2 < \{T_1, C_1,T_2\}$), a fatal event can still be subsequently observed (or censored), unless `non_fatal_censors_fatal = TRUE` is toggled in the main `makeData()` function. This is a global option for all TTEs simulated.

### Semi-Competing Risks Example

We want to generate a data set with TTE endpoints, one fatal and one non-fatal. The fatal event has an average time of 50 in the control group, and 60 in the treatment group. We want an observed event rate of .25 in the control group. 

To solve for the needed log hazard ratio, we solve $\frac{1}{60} = \frac{1}{50} \cdot HR \to \text{HR} = .833$. To solve for the censoring rate, we solve $\frac{1/50}{0.25} - \frac{1}{50} = 1/ 16.667$.

For the second endpoint, we want a mean time of 35 for control group and 45 for treatment group, with an observed event rate of .45 for the control group. To solve for the needed log hazard ratio, we solve $\frac{1}{45} = \frac{1}{35} \cdot HR \to \text{HR} = 0.778$. To solve for the censoring rate, we solve $\frac{1/35}{0.45} - \frac{1}{35} = 1/ 28.64$.

We assume a correlation of .2 between the endpoints.

```{r}
tte_fatal <-  list(
      endpoint_type  = "tte",
      baseline_rate  = 1/50,        
      trt_effect     = log(0.8333),   
      censoring_rate = 1/16.667,        
      fatal_event    = TRUE   
    ) 

tte_nonfatal <- list(
      endpoint_type  = "tte",
      baseline_rate  = 1/35,        
      trt_effect     = log(0.778),   
      censoring_rate = 1/28.64,        
      fatal_event    = FALSE 
    ) 
```


```{r}
small_cor <- corr_make(
  num_endpoints = 2,
  values = rbind( c(1,2, 0.20)) # correlation
)

tte_ex <- makeData(
  correlation_matrix    = small_cor,
  sample_size_per_group = 5000,
  SEED = 321,
  endpoint_details = list( tte_fatal, tte_nonfatal),
  non_fatal_censors_fatal = FALSE
)
tte_ex
```

And we can confirm our inputs:

<details>
<summary>Show summary(tte_ex)</summary>

```{r}
#| echo: false
#| results: asis
x <- paste(capture.output(summary(tte_ex)), collapse = "\n")
cat("<pre>", x, "</pre>", sep = "\n")
```
</details> 

We emphasize once again that censoring and the presence of competing risks can alter correlation (estimated vs expected) and marginal distribution parameters. As mentioned before, we recommend first generating data with no censoring or fatal events to confirm the overall structure before introducing those elements.

In the `summary()` output for simulated TTE endpoints, the column `exp_rate` reports the **arm-specific exponential maximum likelihood estimate (MLE)** of the event rate, computed from the observed times and event indicators after censoring/fatal-event logic has been applied. For a given arm, this is

$$
\widehat{\lambda}_{\text{exp}} = \frac{\sum_i \delta_i}{\sum_i X_i},
$$

where $X_i$ is the observed follow-up time for subject $i$ (possibly censored), and $\delta_i\in\{0,1\}$ is the event indicator (1 = event observed, 0 = censored).

This is the standard MLE for an exponential survival model with right censoring. In the summary output:

- `obs_event_rate` is the observed proportion of subjects with an event (i.e., $\bar\delta$),
- `exp_rate` is the exponential rate estimate $\sum \delta_i / \sum X_i$,
- `est_trt_logHR` is the treatment log-hazard ratio estimated from a Cox model fit.

These quantities are related but not identical. In particular, `exp_rate` is a **descriptive arm-level estimator** under an exponential model assumption, while `est_trt_logHR` comes from a **semi-parametric Cox model**.

Because censoring, administrative censoring, and fatal-event truncation modify the observed follow-up times and event indicators, the estimated `exp_rate` may differ from the user-specified `baseline_rate` (and implied treatment-group rates), especially in small samples or when censoring is heavy.

In this example, the arm-specific exp_rate values are close to the data-generating exponential event rates (approximately $1 / 50,1 / 60,1 / 35$, and $1 / 45$ ), which is expected here because the sample size is large and censoring is independent; even though censoring changes the observed event proportion, the exponential MLE $\sum \delta_i / \sum X_i$ remains a consistent estimator of the event hazard under the exponential model.


# Stochastic enrollment, Follow-Up, and Trial Duration Rules

`endpoints` also allows for a detailed trial-calendar interface through the `enrollment_details`, `followup_details`, and `trial_end_details` arguments. These features allow users to simulate more realistic trials, including staggered enrollment/recruitment, calendar-based follow-up limits, and event-driven final analyses.

For a full tutorial, see:

```{r,eval=FALSE}
vignette("enrollment_guide", package = "endpoints")
```

Briefly, these arguments support:

- stochastic enrollment with `enrollment_distribution = "none"`, `"exponential"`, or `"piecewise"`
- trial-ending rules through `trial_end_details`, including fixed calendar end, last-patient-minimum-follow-up end, or event-driven end
- subject-level follow-up limits through `followup_details`, where `min_followup` defines a required minimum follow-up window for late enrollees and `max_followup` caps how long any subject can be observed
- realized calendar outputs through the returned `enrollTime`, `availableFollowup`, and `sim$meta$trial_calendar`

Below we show two compact TTE examples. The first uses a
last-patient-minimum-follow-up design, while the second uses an event-driven
stopping rule.

## Example: last-patient-minimum-follow-up design

In this example, subjects enroll gradually, each subject can contribute at
most 36 months of follow-up, and the final analysis occurs only after the last
randomized subject has at least 24 months of follow-up.

```{r}
tte_calendar_ep <- list(
  endpoint_type = "tte",
  baseline_rate = rate_from_prob(
    target_prob = 0.30,
    mode = "admin",
    admin_time = 12
  ),
  trt_effect = log(0.70),
  censoring_rate = 0.015
)

sim_lpmf <- makeData(
  correlation_matrix = NULL,
  sample_size_per_group = 100,
  SEED = 42,
  endpoint_details = list(tte_calendar_ep),
  enrollment_details = list(
    enrollment_distribution = "exponential",
    enrollment_exponential_rate = 8
  ),
  followup_details = list(
    min_followup = 24,
    max_followup = 36
  ),
  trial_end_details = list(
    type = "last_patient_min_followup"
  )
)

dat_lpmf <- as.data.frame(sim_lpmf)
tc_lpmf <- sim_lpmf$meta$trial_calendar

tibble(
  trial_end_time = tc_lpmf$trial_end_time,
  trial_end_reason = tc_lpmf$trial_end_reason,
  mean_available_followup = mean(dat_lpmf$availableFollowup),
  observed_events = sum(dat_lpmf$Status_1)
) %>%
  knitr::kable(
    digits = 2,
    caption = "Last-patient-minimum-follow-up example"
  )
```

Here the analysis time depends on when the final subject enrolls. Earlier
subjects may contribute more follow-up, but each subject remains capped at
`max_followup = 36`.

## Example: event-driven trial end

We now switch to an event-driven design. In this case, enrollment is still
stochastic, but the trial ends once the chosen TTE endpoint reaches a target
number of observed events.

```{r}
sim_event_driven <- makeData(
  correlation_matrix = NULL,
  sample_size_per_group = 120,
  SEED = 99,
  endpoint_details = list(tte_calendar_ep),
  enrollment_details = list(
    enrollment_distribution = "piecewise",
    piecewise_enrollment_cutpoints = c(0, 6, 12, 24),
    piecewise_enrollment_rates = c(4, 8, 12)
  ),
  followup_details = list(
    max_followup = 30
  ),
  trial_end_details = list(
    type = "event_driven",
    event_endpoint = "TTE_1",
    target_events = 70
  )
)

dat_event_driven <- as.data.frame(sim_event_driven)
tc_event_driven <- sim_event_driven$meta$trial_calendar

tibble(
  trial_end_time = tc_event_driven$trial_end_time,
  trial_end_reason = tc_event_driven$trial_end_reason,
  event_target_reached = tc_event_driven$event_target_reached,
  n_events_at_end = tc_event_driven$n_events_at_end,
  mean_available_followup = mean(dat_event_driven$availableFollowup)
) %>%
  knitr::kable(
    digits = 2,
    caption = "Event-driven example"
  )
```

The realized calendar information is stored in `sim_event_driven$meta$trial_calendar`.
This metadata records why the trial stopped, when it stopped, and how many
events were observed at the realized analysis time.

# References

Cario, M. C., & Nelson, B. L. (1997). *Modeling and generating random vectors with arbitrary marginal distributions and correlation matrix* (pp. 1-19). Technical Report, Department of Industrial Engineering and Management Sciences, Northwestern University, Evanston, Illinois.





```{r,echo=FALSE}
knitr::knit_exit()
```
